Place of death in NRS COVID-19 deaths

### 2021-02-22 covid_nrs_place_of_death.Rmd
modified the scripts not to pass params to .Rmd but instead to load the latest data csv instead

---
title: "Place of death in NRS COVID-19 deaths"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
params:
  data_sources: '`r tibble()`'
  historical_data_sources: '`r tibble()`'
  case_trends_data_sources: '`r tibble()`'
---

@@

```{r}
endpoint <- "http://statistics.gov.scot/sparql.json"
endpoint <- "http://statistics.gov.scot/sparql"
```

@@

# Set up output directory

@@

```{r}
# "unpack" the data source tables
table_of_data_sources <- params$data_sources
table_of_historical_data_sources <- params$historical_data_sources
table_of_case_trends_data_sources <- params$case_trends_data_sources

@@

ods_all_datasets() %>% View()
ods_structure("deaths")  # can we import all deaths-related data this way?
ods_structure("coronavirus-covid-19-management-information")

@@

total_deaths_hb %>% filter(year==2020 & str_starts(place_of_death, pattern="Home")) %>%
  left_join(cumulative_case_2020_based_grouping_hb, by="ref_area") %>%
  ggplot(data=., aes(x=cumulative_positive, y=deaths_all_causes, label=ref_area)) +
  geom_point() +
  geom_label_repel() +
  theme_minimal() +
  NULL

# k means clustering
hb_clustered <- kmeans(cumulative_case_2020_based_grouping_hb %>% filter(ref_area!="Scotland") %>% select(2:3), centers = 4, nstart = 2500)
bind_cols(cumulative_case_2020_based_grouping_hb %>% filter(ref_area!="Scotland"), cluster=hb_clustered$cluster) %>%
  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, fill=cluster, colour=cluster)) +
    geom_point() +
    geom_label_repel() +
    theme_minimal() +
    NULL

@@

```{r}
## this is fairly ad-hoc... but it works!
table_comparing_rates <-
  bind_rows(
    ## all available data
    merged_data_overall %>%
      filter(age == "all" &
               sex == "all" &
               str_starts(place_of_death, "Home")) %>% group_by(year, week_number) %>%
      summarise(
        deaths_all_causes = sum(deaths_all_causes),
        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        .groups = "drop"
      ) %>% left_join(
        past_deaths_overall %>% filter(str_starts(place_of_death, "Home") &
                                         year != "average") %>% group_by(week_number) %>% summarise(
                                           average_weekly_deaths = sum(number_of_deaths) / 5,
                                           .groups = "drop"
                                         ),
        by = "week_number"
      ) %>% summarise(
        period = "all available data",
        deaths_all_causes = sum(deaths_all_causes),
        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        past_annual_deaths = sum(average_weekly_deaths)
      )
    ,
    ## 2020 only
    bind_cols(
      merged_data_overall %>%
        filter(
          age == "all" &
            sex == "all" & str_starts(place_of_death, "Home") & year == 2020
        ) %>%
        summarise(
          period = "2020",
          deaths_all_causes = sum(deaths_all_causes),
          deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
          deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE)
        ),
      past_deaths_overall %>% filter(str_starts(place_of_death, "Home") &
                                       year != "average") %>% summarise(past_annual_deaths = sum(number_of_deaths) / 5)
    ),
    ## 2020 from week 12 onwards (including week 12) - pandemic part of 2020
    bind_cols(
      merged_data_overall %>%
        filter(
          age == "all" &
            sex == "all" &
            str_starts(place_of_death, "Home") &
            year == 2020 & week_number >= 12
        ) %>%
        summarise(
          period = "2020, from wk12",
          deaths_all_causes = sum(deaths_all_causes),
          deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
          deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE)
        ),
      past_deaths_overall %>% filter(
        week_number >= 12 &
          str_starts(place_of_death, "Home") &
          year != "average"
      ) %>% summarise(past_annual_deaths = sum(number_of_deaths) / 5)
    )
  ) %>%
  mutate(
    ratio_deaths_all_causes_to_historical = deaths_all_causes / past_annual_deaths,
    ratio_non_covid_deaths_to_historical = deaths_non_covid / past_annual_deaths,
    proportion_covid_deaths = deaths_covid_related / deaths_all_causes
  )

@@

source_nrs_text <- "Source: National Records of Scotland"

@@

# figure_home_deaths_and_cases_scotland <-
  merged_deaths_overall %>%
  filter(sex=="all" & age=="all" & str_starts(place_of_death, "Home")) %>%
  left_join(weekly_covid_cases_overall, by=c("year","week_number","date_w_c","week_number_run_over","ref_area")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  geom_line(aes(y = weekly_positive * ratio_max_weekly_deaths_to_cases), colour="purple") +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  scale_y_continuous(sec.axis = sec_axis(trans = ~./ratio_max_weekly_deaths_to_cases, name = "Weekly deaths")) +
  # ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  annotate(geom = "text", label = total_deaths_at_home_in_period, x = 25, y = 575, hjust=0) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(dashed_line_historical_range_text,vertical_line_text, death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

>>>>

---
title: "Place of death in NRS COVID-19 deaths"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

@@

```{r}
# endpoint <- "http://statistics.gov.scot/sparql.json"
endpoint <- "http://statistics.gov.scot/sparql"
```

@@

# Load info about latest data

A separate script checks for updated data and downloads it, also updating these files with the locations of the downloaded data:

```{r}
file_data_sources <- "./table_of_data_sources.csv"
file_historical_data_sources <- "./table_of_historical_data_sources.csv"
file_case_trends_data_sources <- "./table_of_case_trends_data_sources.csv"

table_of_data_sources <- read_csv(file = file_data_sources)
table_of_historical_data_sources <- read_csv(file = file_historical_data_sources)
table_of_case_trends_data_sources <- read_csv(file = file_case_trends_data_sources)

## Set up output directory

@@

```{r}
# TODO: remove this section, params not used anymore
# # "unpack" the data source tables
# table_of_data_sources <- params$data_sources
# table_of_historical_data_sources <- params$historical_data_sources
# table_of_case_trends_data_sources <- params$case_trends_data_sources
```

@@

# ods_all_datasets() %>% View()
# ods_structure("deaths")  # can we import all deaths-related data this way?
# ods_structure("coronavirus-covid-19-management-information")

@@

hb_cases_plots_assembled
```

@@

```{r}
## this is fairly ad-hoc... but it works!
table_comparing_rates <-
  bind_rows(
    ## all available data
    merged_deaths_overall %>%
      filter(age == "all" &
               sex == "all" &
               str_starts(place_of_death, "Home")) %>% group_by(year, week_number) %>%
      summarise(
        deaths_all_causes = sum(deaths_all_causes),
        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        .groups = "drop"
      ) %>% left_join(
        past_deaths_overall %>% filter(str_starts(place_of_death, "Home") &
                                         year != "average") %>% group_by(week_number) %>% summarise(
                                           average_weekly_deaths = sum(number_of_deaths) / 5,
                                           .groups = "drop"
                                         ),
        by = "week_number"
      ) %>% summarise(
        period = "all available data",
        deaths_all_causes = sum(deaths_all_causes),
        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        past_annual_deaths = sum(average_weekly_deaths)
      )
    ,
    ## 2020 only
    bind_cols(
      merged_deaths_overall %>%
        filter(
          age == "all" &
            sex == "all" & str_starts(place_of_death, "Home") & year == 2020
        ) %>%
        summarise(
          period = "2020",
          deaths_all_causes = sum(deaths_all_causes),
          deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
          deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE)
        ),
      past_deaths_overall %>% filter(str_starts(place_of_death, "Home") &
                                       year != "average") %>% summarise(past_annual_deaths = sum(number_of_deaths) / 5)
    ),
    ## 2020 from week 12 onwards (including week 12) - pandemic part of 2020
    bind_cols(
      merged_deaths_overall %>%
        filter(
          age == "all" &
            sex == "all" &
            str_starts(place_of_death, "Home") &
            year == 2020 & week_number >= 12
        ) %>%
        summarise(
          period = "2020, from wk12",
          deaths_all_causes = sum(deaths_all_causes),
          deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
          deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE)
        ),
      past_deaths_overall %>% filter(
        week_number >= 12 &
          str_starts(place_of_death, "Home") &
          year != "average"
      ) %>% summarise(past_annual_deaths = sum(number_of_deaths) / 5)
    )
  ) %>%
  mutate(
    ratio_deaths_all_causes_to_historical = deaths_all_causes / past_annual_deaths,
    ratio_non_covid_deaths_to_historical = deaths_non_covid / past_annual_deaths,
    proportion_covid_deaths = deaths_covid_related / deaths_all_causes
  )

@@

source_nrs_text <- "Source: National Records of Scotland"
source_phs_text <- "Source: Public Health Scotland"

@@

# figure_home_deaths_and_cases_scotland <-
  merged_deaths_overall %>%
  filter(sex=="all" & age=="all" & str_starts(place_of_death, "Home")) %>%
  left_join(weekly_covid_cases_overall, by=c("year","week_number","date_w_c","week_number_run_over","ref_area")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  geom_line(aes(y = weekly_positive * ratio_max_weekly_deaths_to_cases), colour="purple") +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  scale_y_continuous(sec.axis = sec_axis(trans = ~./ratio_max_weekly_deaths_to_cases, name = "Weekly cases")) +
  # ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  annotate(geom = "text", label = total_deaths_at_home_in_period, x = 25, y = 575, hjust=0) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(dashed_line_historical_range_text,vertical_line_text, death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

  merged_deaths_overall %>%
  filter(sex=="all" & age=="all" & str_starts(place_of_death, "Home")) %>%
  left_join(weekly_covid_cases_overall, by=c("year","week_number","date_w_c","week_number_run_over","ref_area")) %>%
  ggplot(aes(x=week_number_run_over, y=weekly_positive)) +
  geom_line() +
  theme_minimal() +
  scale_y_continuous(labels = function(x) format(x, big.mark=",")) +
  labs(
    x = "Week number", y = "COVID-19 positive cases (N)",
    subtitle = date_range_text,
    caption = paste0(c(vertical_line_text, source_phs_text), collapse="\n")
  )
